Calendar-based graphics for visualising people's daily schedules

### 2017-07-05 main.Rmd
intro roughly done

---
title: "Calendar-based graphics for visualising people's daily schedules"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Di
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
fontsize: 11pt
papersize: a4
fontfamily: mathpazo
blind: false
cover: true
wpnumber: no/yr
jelcodes: C10,C14,C22
bibliography: ["references.bib", "packages.bib"]
biblio-style: authoryear-comp
abstract: "A brief summary of our ideas"
keywords: "blah, blah"
subparagraph: true
toc: false
output: 
  bookdown::pdf_document2:
    template: monashwp.tex
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    number_sections: yes
    citation_package: biblatex
---

@@

Calendar-based visualisation proves useful in unfolding human-related activities in temporal context. For example, @van_wijk_cluster_1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It effectively contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: `ggTimeSeries` [@R-ggTimeSeries] and `ggcal` [@R-ggcal]. However, their technique is too constrained to colour-encoding graphics. We shall extend this calendar-based arrangement using linear algebra tools from a single type of heatmap to a broader range of glyphs, from univariate to multivariate cases, from one calendar format to three formats. The proposed algorithm has been developed in the `sugrrants` package [@R-sugrrants] using R [@R-base].

This work was originally motivated by studying hourly foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed across downtown Melbourne since 2009 in order to capture the pulse of city life. (FIGURE ?: weekday at multiple sensors) Figure (?) lends itself to addressing some of the issues: (1) to simultaneously exhibit multiple seasons including time of day, day of week, and day of year, (2) to visualise long historical time series data in an information-dense way at the overview level, (3) to compare and contrast multiple time series on a common scale, (4) to quickly look up the date when there are unexpected events happening without interactivity.

# Construction

@@

@@

\begin{equation}
  \begin{aligned}
  i &= \lceil (g \mod 35) / 7\rceil, \\
  j &= g \mod 7. (\#eq:grid)
  \end{aligned}
\end{equation}

Figure \@ref(fig:month-diagram) illustrates this $(i,j)$ layout for a month where $k=5$.

@@

To create the layout for a full year, $(m, n)$ denotes the position of the month arranged in the plot, where $1 \le m \le M$ and $1 \le n \le N$. Between each month requires some small amount of white space, label this $b$. Figure \@ref(fig:year-diagram) illustrates this  layout.

@@

The abbreviated month labels located on the top left are obtained through $(\min{(x)}, \max{(y)})$ for every $m$ and $n$. The weekday labels are uniformly placed on the bottom of the whole canvas, that is $\min{(y)}$.

# The `frame_calendar` function

```{r args, eval = FALSE, echo = TRUE}
frame_calendar(data, x, y, date, calendar = "monthly", dir = "h",
  sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

# Discussion

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Di Cook
    affiliation: Monash Univeristy
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash Univeristy
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [calendar-based visualisation, temporal-context data]
  plain:     [calendar-based visualisation, temporal-context data]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. Since 2009, there have been 43 installed sensors counting pedestrians every hour across downtown Melbourne. The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. (FIGURE ?: weekday at multiple sensors) Figure (?) lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

1. Variations primarily result from multiple time scales including time of day, day of week, and day of year (such as public holiday and recurred events).
2. Since the data are often collected at daily frequency or a time scale more frequent than daily, they typically involve large number of observations spanning over a long time period.
3. Measurements of a single type are made at multiple locations at a given time point, which creates a need in comparing and contrasting between locations.

Calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context (cite). For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It remarkably contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week considered as the smallest time window. Time of day, which serves as one of the most important aspect in explaining variations arising from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colours encoded as quantities, it may becomes less perceivable of the actual variations or behaviours from the viewer's perspective.

We propose a new algorithm via linear algebra tools to go beyond the calendar-based heatmap. The approach is developed with these conditions in mind: (1) to make time of day present in addition to the existing temporal components such as day of week and day of year, (2) to add line graphs along with other types of glyphs to the visual toolkit, (3) to enable an overlaying plot consisting of multiple time series. The proposed algorithm has been implemented as the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

The remainder of the paper is organised as follows. Section \ref{sec:algorithm} describes the \code{frame_calendar} function and its construction in detail. Section \ref{sec:examples} presents examples using a range of geometric objects. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

# Construction 
\label{sec:algorithm}

@@

```
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

@@

\begin{equation}
  \begin{aligned}
  i &= \lceil (g \mod 35) / 7\rceil, \\
  j &= g \mod 7. \label{eq:grid}
  \end{aligned}
\end{equation}

Figure \ref{fig:month-diagram} illustrates this $(i,j)$ layout for a month where $k=5$.

@@

To create the layout for a full year, $(m, n)$ denotes the position of the month arranged in the plot, where $1 \le m \le M$ and $1 \le n \le N$. Between each month requires some small amount of white space, label this $b$. Figure \ref{fig:year-diagram} illustrates this  layout.

@@

The abbreviated month labels located on the top left are obtained through $(\min{(x)}, \max{(y)})$ for every $m$ and $n$. The weekday texts with a single letter are uniformly placed on the bottom of the whole canvas, that is $\min{(y)}$.

# Examples
\label{sec:examples}

# Discussion
\label{sec:discussion}

### 2017-07-05 main.tex
intro roughly done

Calendar-based visualisation turns out to be a useful tool in unfolding
human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It
effectively contrasts weekdays and weekends, highlights public holiday,
and presents other known seasons like school vacations, all of which
have influence over the turn-outs in the office. The calendar-based
heatmap was implemented in a couple of R packages: \texttt{ggTimeSeries}
\citep{R-ggTimeSeries} and \texttt{ggcal} \citep{R-ggcal}. However,
their techniques are too constrained to colour-encoding graphics. We
shall extend this calendar-based arrangement using linear algebra tools
from a single type of heatmap to a broader range of glyphs, from
univariate to multivariate cases, from one calendar format to three
formats. The proposed algorithm has been developed in the
\texttt{sugrrants} package \citep{R-sugrrants} using R \citep{R-base}.

This work was originally motivated by studying hourly foot traffic in
the city of Melbourne \citep{ped}. There have been 43 sensors installed
across downtown Melbourne since 2009 in order to capture the pulse of
city life. (FIGURE ?: weekday at multiple sensors) Figure (?) lends
itself to addressing some of the issues:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  to simultaneously exhibit multiple seasons including time of day, day
  of week, and day of year
\item
  to visualise long historical time series data in an information-dense
  way at the overview level
\item
  to compare and contrast multiple time series on a common scale
\item
  to quickly look up the date when there are unexpected events happening
  without interactivity.
\end{enumerate}

\section{The function and its
construction}\label{the-function-and-its-construction}

@@

@@

>>>>

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. Since 2009, there have been 43 installed
sensors counting pedestrians every hour across downtown Melbourne. The
dataset can shed lights into understanding people's daily schedules, or
assisting administration and business planning. (FIGURE ?: weekday at
multiple sensors) Figure (?) lends itself to a number of issues that
make exploratory data visualisation challenging in many temporal-context
applications involving human behaviours:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations primarily result from multiple time scales including time
  of day, day of week, and day of year (such as public holiday and
  recurred events).
\item
  Since the data are often collected at daily frequency or a time scale
  more frequent than daily, they typically involve large number of
  observations spanning over a long time period.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates a need in comparing and contrasting
  between locations.
\end{enumerate}

Calendar-based graphics turn out to be a useful tool in unfolding
human-related activities in temporal context (cite). For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It remarkably
contrasts weekdays and weekends, highlights public holiday, and presents
other known seasons like school vacations, all of which have influence
over the turn-outs in the office. The calendar-based heatmap was
implemented in a couple of R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are too constrained to colour-encoding graphics and day of
week considered as the smallest time window. Time of day, which serves
as one of the most important aspect in explaining variations arising
from pedestrian sensor data, will be neglected through daily
aggregation. Additionally if simply using colours encoded as quantities,
it may becomes less perceivable of the actual variations or behaviours
from the viewer's perspective.

We propose a new algorithm via linear algebra tools to go beyond the
calendar-based heatmap. The approach is developed with these conditions
in mind: (1) to make time of day present in addition to the existing
temporal components such as day of week and day of year, (2) to add line
graphs along with other types of glyphs to the visual toolkit, (3) to
enable an overlaying plot consisting of multiple time series. The
proposed algorithm has been implemented as the \code{frame_calendar}
function in the \pkg{sugrrants} package \citep{R-sugrrants} using
\proglang{R} \citep{R-base}.

The remainder of the paper is organised as follows. Section
\ref{sec:algorithm} describes the \code{frame_calendar} function and its
construction in detail. Section \ref{sec:examples} presents examples
using a range of geometric objects. Section \ref{sec:discussion}
discusses the advantages and disadvantages of the method.

\section{Construction}\label{construction}

\label{sec:algorithm}

@@

\label{sec:examples}

@@

\label{sec:discussion}
